---
title: "1. Getting Started with keRnel (Beginner)"
subtitle: "Every base kernel, printing, evaluation, missing data, and basic composition"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{1. Getting Started with keRnel (Beginner)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(keRnel)
```

## What is a covariance kernel?

A covariance kernel is a function `k(x1, x2)` that measures how "similar"
two input points are. Gaussian Process models use a kernel to build a
covariance matrix over a set of points: how correlated two outputs are
depends on how similar their inputs are, according to the kernel.

`keRnel` provides a large catalogue of kernels, all sharing the exact same
interface:

```{r}
k <- se_kernel(length_scale = 1)
k
```

Every kernel is built by a constructor function (`se_kernel()` above), and
every kernel can be evaluated on a set of points with `evaluate()`:

```{r}
evaluate(k, c(0, 1, 2))
```

The result is the 3x3 covariance matrix between the three points `0`, `1`,
and `2`. Notice the diagonal is always `1` here: a point is always
maximally similar to itself for this kernel.

If you only pass one set of points, `evaluate()` computes the matrix of a
point set against itself (the common "auto-covariance" case for fitting a
GP). You can also pass two different sets of points to get a *cross*-
covariance matrix:

```{r}
evaluate(k, c(0, 1), c(0, 10, 20))
```

The result has 2 rows (length of the first argument) and 3 columns (length
of the second).

## Printing a kernel

Every kernel prints its hyperparameters in their natural, constrained
space (e.g. `length_scale` is always shown positive, never in whatever
internal representation is used for optimisation):

```{r}
print(k)
matern32_kernel(length_scale = 2.5)
periodic_kernel(length_scale = 1, period = 7)
```

## Multi-dimensional inputs

A bare numeric vector like `c(0, 1, 2)` is always interpreted as **N points
of 1 dimension** -- never as a single point of N dimensions. For points
with more than one dimension, pass a matrix instead, one row per point:

```{r}
X <- matrix(c(
  0, 0,
  1, 0,
  0, 1
), ncol = 2, byrow = TRUE)
X
evaluate(k, X)
```

::: {.alert .alert-info}
**Sharp bit**: if you want to evaluate a kernel at a *single*
multi-dimensional point, you must still pass it as a 1-row matrix
(`matrix(c(1, 2), nrow = 1)`), not as a bare vector `c(1, 2)` -- the latter
is read as 2 points of 1 dimension each.
:::

## The full catalogue of base kernels

Every base kernel constructor validates its hyperparameters and gives a
sensible error if you pass something invalid (e.g. a non-positive
`length_scale`):

```{r, error = TRUE}
keRnel::se_kernel(length_scale = 1)
```

### Stationary kernels (depend only on the distance between points)

```{r}
# Squared Exponential / RBF (rbf_kernel is an alias)
evaluate(se_kernel(length_scale = 1), c(0, 1, 2))

# Matern family (increasing smoothness: 1/2, 3/2, 5/2)
evaluate(matern12_kernel(length_scale = 1), c(0, 1, 2))
evaluate(matern32_kernel(length_scale = 1), c(0, 1, 2))
evaluate(matern52_kernel(length_scale = 1), c(0, 1, 2))

# Periodic
evaluate(periodic_kernel(length_scale = 1, period = 2), c(0, 1, 2))

# Rational Quadratic
evaluate(rational_quadratic_kernel(length_scale = 1, alpha = 2), c(0, 1, 2))

# White noise: non-zero only on the diagonal (x1 == x2)
evaluate(white_noise_kernel(noise = 0.1), c(0, 1, 2))

# Feature kernel: a more advanced 2-channel kernel, see the expert vignette
# for its use with block_kernel()
evaluate(feature_kernel(length_scale = 1, length_scale_u = 1, variance = 1), c(0, 1, 2))
```

### Dot-product kernels (depend on the dot product between points)

```{r}
# Linear: passes through the origin
evaluate(linear_kernel(slope_var = 1), c(0, 1, 2))

# Affine: passes through (offset, 0) instead of the origin
evaluate(affine_kernel(slope_var = 1, offset = 1), c(0, 1, 2))

# Polynomial
evaluate(polynomial_kernel(degree = 2, gamma = 1, constant = 1), c(0, 1, 2))

# Sigmoid (hyperbolic tangent)
evaluate(sigmoid_kernel(alpha = 0.5, constant = 0), c(0, 1, 2))
```

### Other kernels (no notion of distance at all)

```{r}
# Constant: the same value everywhere, can be negative
evaluate(constant_kernel(value = 2), c(0, 1, 2))

# Variance: the same strictly positive value everywhere, typically used to
# scale another kernel (see "Combining kernels" below)
evaluate(variance_kernel(variance = 3), c(0, 1, 2))
```

## Combining kernels with `+` and `*`

Kernels combine just like numbers, using `+` and `*`:

```{r}
k_sum <- se_kernel(length_scale = 1) + white_noise_kernel(noise = 0.1)
k_sum
evaluate(k_sum, c(0, 1, 2))
```

```{r}
k_product <- variance_kernel(variance = 4) * se_kernel(length_scale = 1)
k_product
evaluate(k_product, c(0, 1, 2))
```

A bare number on either side of `+`/`*` is automatically treated as a
`constant_kernel()`:

```{r}
k_plus_constant <- se_kernel(length_scale = 1) + 0.5
k_plus_constant
```

You can chain as many `+`/`*` as you like:

```{r}
k_chain <- se_kernel(length_scale = 1) * variance_kernel(variance = 2) +
  white_noise_kernel(noise = 0.05)
k_chain
```

::: {.alert .alert-info}
**Sharp bit**: unlike many other GP libraries, `keRnel` deliberately does
**not** support `-` (subtraction) or unary negation (`-k`) for kernels --
subtracting covariance kernels is rarely mathematically meaningful (it can
easily break positive semi-definiteness) and was judged unnecessary.
:::

```{r, error = TRUE}
se_kernel(length_scale = 1) - white_noise_kernel(noise = 0.1)
```

## Missing data: the default behaviour

Real datasets often have missing values. By default, a missing input point
(`NA`) makes the corresponding row/column of the result `NA` too:

```{r}
evaluate(se_kernel(length_scale = 1), c(0, NA, 2))
```

This is deliberately the safest default: it forces you to notice and
decide what to do about missing data, rather than silently producing a
matrix that looks fine but secretly is not invertible. The intermediate
vignette (`vignette("b-intermediate")`) shows `na_action = "mask"`, an
alternative that keeps the matrix invertible.

## Where to go next

- `vignette("b-intermediate")`: hyperparameter introspection
  (`get_params()`), immutable updates (`kupdate()`), freezing parameters
  for optimisation (`freeze()`), parametrisations, custom distance
  functions, `active_dims_kernel()`, `ard_kernel()`, `sample_hps()`, and
  writing your own kernel.
- `vignette("c-expert")`: computation engines, performance internals,
  `batch_kernel()`, `input_specific_param_kernel()`, `block_kernel()`, and
  `block_diag_kernel()`.
