---
title: "2. Customizing and Fitting Kernels (Intermediate)"
subtitle: "Hyperparameters, immutable updates, freezing, parametrisations, distances, ARD, priors, and custom kernels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{2. Customizing and Fitting Kernels (Intermediate)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(keRnel)
```

This vignette assumes you have read `vignette("a-beginner")`. It covers
everything you need to fit a kernel's hyperparameters to data: introspecting
them, updating them immutably, excluding some from optimisation, and
controlling exactly how each one is constrained.

## Inspecting hyperparameters: `get_params()`

Every kernel stores its hyperparameters internally in an unconstrained
("wrapped") space suitable for a generic optimiser, but exposes them in
their natural, constrained space via `get_params()`:

```{r}
k <- se_kernel(length_scale = 2.5) + se_kernel(length_scale = 1.5)
get_params(k)
```

`get_params()` recurses through any composition of kernels, returning every
hyperparameter found anywhere in the tree:

```{r}
k_composite <- se_kernel(length_scale = 1) + white_noise_kernel(noise = 0.1)
get_params(k_composite)
```

## Immutable updates: `kupdate()`

Kernels are immutable: there is no way to change a kernel "in place".
Instead, `kupdate()` returns a **new** kernel with the requested
hyperparameter(s) changed, leaving the original untouched:

```{r}
k1 <- se_kernel(length_scale = 1)
k2 <- kupdate(k1, length_scale = 5)

get_params(k1) # unchanged
get_params(k2) # updated
```

`kupdate()` is a single generic that works on *any* kernel, however deeply
composed, by finding the hyperparameter wherever it occurs:

```{r}
k3 <- se_kernel(length_scale = 1) * se_kernel(length_scale = 1)
k4 <- kupdate(k3, length_scale = 9) # updates *both* occurrences
get_params(k4)
```

A typo is caught immediately, rather than silently doing nothing:

```{r, error = TRUE}
kupdate(k1, length_scael = 5)
```

::: {.alert .alert-info}
`kupdate()` works the same way on every kernel type because R's S3 dispatch
picks the right update logic for sums, products, wrappers, and base kernels
alike, so there's a single generic to remember rather than a method per
kernel class. It is named `kupdate()` rather than `replace()` because
`base::replace()` already exists in R (and is not an S3 generic), so
overriding it would shadow a base function.
:::

## Freezing parameters: `freeze()`/`unfreeze()`

When fitting a GP, you sometimes want to hold a hyperparameter fixed (not
let the optimiser touch it) while still using it in every computation.
`freeze()` marks a hyperparameter as excluded from `get_trainable_params()`
(the vector an optimiser should actually act on), without removing it from
`get_params()` (the *full* picture of the kernel's current configuration):

```{r}
k <- se_kernel(length_scale = 2) + white_noise_kernel(noise = 0.1)
k_frozen <- freeze(k, "noise")

get_params(k_frozen) # both still listed
get_trainable_params(k_frozen) # noise excluded
```

`unfreeze()` reverses it:

```{r}
k_unfrozen <- unfreeze(k_frozen, "noise")
get_trainable_params(k_unfrozen)
```

A frozen kernel still computes normally -- `freeze()` only affects which
values an optimiser sees, not the kernel's behaviour:

```{r}
evaluate(k, c(0, 1)) - evaluate(k_frozen, c(0, 1)) # identical
```

## Parametrisations: how a hyperparameter is constrained

Every hyperparameter is associated with a **parametrisation**: an object
that knows how to map the value between its natural (constrained) space and
an unconstrained space suitable for a generic optimiser. The default for a
strictly positive hyperparameter (like `length_scale`) is
`log_exp_parametrisation()`:

```{r}
k <- se_kernel(length_scale = 1)
k$length_scale$parametrisation
```

### `log_exp_parametrisation()`: the default for positive values

Maps via `log()`/`exp()`. Validation (strict positivity) happens *inside*
`wrap()`, the single point of truth used both at construction and at every
`kupdate()`:

```{r, error = TRUE}
se_kernel(length_scale = -1)
kupdate(se_kernel(length_scale = 1), length_scale = -1)
```

### `softplus_parametrisation()`: an alternative for positive values

A gentler gradient near zero -- swap it in for any positive hyperparameter:

```{r}
k_softplus <- se_kernel(
  length_scale = 1,
  length_scale_parametrisation = softplus_parametrisation()
)
get_params(k_softplus) # same constrained value...
k_softplus$length_scale$wrapped # ...but a different unconstrained representation
k$length_scale$wrapped
```

### `bounded_parametrisation()`: for a value confined to an interval

Useful for a hyperparameter that must stay strictly within `(lower, upper)`,
e.g. a correlation-like quantity:

```{r}
k_bounded <- se_kernel(
  length_scale = 1,
  length_scale_parametrisation = bounded_parametrisation(lower = 0, upper = 10)
)
get_params(k_bounded)
```

```{r, error = TRUE}
kupdate(k_bounded, length_scale = 20) # outside (0, 10)
```

### `identity_parametrisation()`: no constraint at all

For a hyperparameter that may take any real value (this is the default for
`affine_kernel()`'s `offset` and `polynomial_kernel()`'s `constant`, which
are plain unconstrained fields rather than `wrapped_param`s):

```{r}
identity_parametrisation()
wrap(identity_parametrisation(), -3.5)
```

### `parametrisation_chain()`: composing constraints

Chains several parametrisations together, e.g. positive *and* bounded
above:

```{r}
chain <- parametrisation_chain(log_exp_parametrisation())
unwrap(chain, wrap(chain, 2))
```

## Distance functions

Stationary and dot-product kernels delegate to a `distance_function`. Every
kernel has a sensible default (e.g. squared Euclidean for `se_kernel()`),
but you can swap it for any of the alternatives `keRnel` provides:

```{r}
euclidean_distance(c(0, 0), c(3, 4))
manhattan_distance(c(0, 0), c(3, 4))
chebyshev_distance(c(0, 0), c(3, 4))
cosine_distance(c(1, 0), c(0, 1))
minkowski_distance(c(0, 0), c(3, 4), p = 3)
hamming_distance(c(1, 2, 3), c(1, 0, 3))
canberra_distance(c(1, 2), c(2, 1))
```

As a convenience, you can pass a string shortcut instead of the function
itself:

```{r}
k_manhattan <- se_kernel(length_scale = 1, distance_function = "manhattan")
k_manhattan$distance_function
identical(k_manhattan$distance_function, manhattan_distance)
```

You can also write your own distance function -- any function of two
numeric vectors returning a single value works (it just won't benefit from
`keRnel`'s internal vectorised fast paths, see `vignette("c-expert")`):

```{r}
my_distance <- function(x1, x2) sum((x1 - x2)^2) / 2
k_custom_distance <- se_kernel(length_scale = 1, distance_function = my_distance)
evaluate(k_custom_distance, c(0, 1, 2))
```

## Exponential and logarithm wrappers

`exp_kernel()`/`log_kernel()` apply `exp()`/`log()` to another kernel's
result -- e.g. useful to guarantee strict positivity of an otherwise
unconstrained combination:

```{r}
k_exp <- exp_kernel(se_kernel(length_scale = 1))
evaluate(k_exp, c(0, 1, 2))
```

```{r}
identical(
  evaluate(log_kernel(exp_kernel(se_kernel(length_scale = 1))), c(0, 1, 2)),
  evaluate(se_kernel(length_scale = 1), c(0, 1, 2))
)
```

## Combinatorial helper: `kernel_interactions()`

For additive-with-interactions GP models, manually writing out every
sum/product term gets tedious quickly. `kernel_interactions()` builds the
full expansion automatically:

```{r}
k1 <- se_kernel(length_scale = 1)
k2 <- linear_kernel(slope_var = 1)
k3 <- periodic_kernel(length_scale = 1, period = 2)

k_full <- kernel_interactions(list(k1, k2, k3))
k_full # k1 + k2 + k3 + k1*k2 + k1*k3 + k2*k3 + k1*k2*k3
```

Limit the expansion to lower-order interactions with `max_order`:

```{r}
k_pairwise_only <- kernel_interactions(list(k1, k2, k3), max_order = 2)
k_pairwise_only # no triple product k1*k2*k3
```

## Selecting input dimensions: `active_dims_kernel()`

Restrict a kernel to a subset of input dimensions -- useful to build an
additive kernel over different (possibly overlapping) groups of features:

```{r}
X <- matrix(c(
  0, 10,
  1, 11,
  2, 12
), ncol = 2, byrow = TRUE)

k_dim1 <- active_dims_kernel(se_kernel(length_scale = 1), active_dims = 1)
evaluate(k_dim1, X) # only column 1 (0, 1, 2) is seen

k_dim2 <- active_dims_kernel(se_kernel(length_scale = 1), active_dims = 2)
evaluate(k_dim2, X) # only column 2 (10, 11, 12) is seen
```

Combine with `+` for a fully additive kernel across dimensions:

```{r}
k_additive <- active_dims_kernel(se_kernel(length_scale = 1), active_dims = 1) +
  active_dims_kernel(matern32_kernel(length_scale = 1), active_dims = 2)
evaluate(k_additive, X)
```

## Automatic Relevance Determination: `ard_kernel()`

ARD gives every input dimension its own `length_scale`, so the model can
learn that some features matter more than others:

```{r}
k_ard <- ard_kernel(se_kernel, length_scales = c(1, 5, 0.1))
X3 <- matrix(rnorm(9), ncol = 3)
dim(evaluate(k_ard, X3))
```

`ard_kernel()` takes a **constructor** function (not a pre-built instance)
and handles construction and freezing itself, so there is no way to build
it incorrectly:

```{r}
k_ard$subkernels[[1]]$length_scale # always exactly 1...
k_ard$subkernels[[1]]$length_scale$frozen # ...and always frozen
get_trainable_params(k_ard) # only length_scales is trainable
```

Extra constructor arguments pass through via `...`:

```{r}
k_ard_periodic <- ard_kernel(periodic_kernel, length_scales = c(1, 2), period = 5)
k_ard_periodic$subkernels[[1]]$period
```

## Random initialisation: `sample_hps()`

For multi-start optimisation, you often want several random initial
configurations. `sample_hps()` resamples named hyperparameters from a
prior, anywhere they occur in the kernel tree:

```{r}
set.seed(123)
k <- se_kernel(length_scale = 1) + white_noise_kernel(noise = 0.1)

k_sampled <- sample_hps(k, priors = list(
  length_scale = c(0.1, 10), # c(min, max) for the default uniform prior
  noise = c(0.001, 1)
))
get_params(k_sampled)
```

Normal priors are supported via `distribution = stats::rnorm` (then
`priors` holds `c(mean, sd)` instead of `c(min, max)`):

```{r}
k_sampled_normal <- sample_hps(
  se_kernel(length_scale = 1),
  priors = list(length_scale = c(2, 0.5)),
  distribution = stats::rnorm
)
get_params(k_sampled_normal)
```

Frozen hyperparameters are skipped, consistent with `get_trainable_params()`:

```{r}
k_frozen <- freeze(se_kernel(length_scale = 1), "length_scale")
identical(
  get_params(k_frozen),
  get_params(sample_hps(k_frozen, priors = list(length_scale = c(0.1, 10))))
)
```

::: {.alert .alert-info}
`sample_hps()` draws from R's ambient RNG state, so call `set.seed()`
beforehand for reproducibility, the standard R convention.
:::

## NA handling, revisited: `na_action = "mask"`

`vignette("a-beginner")` showed the default `na_action = "propagate"`. For
GP fitting, you often need the covariance matrix to stay invertible despite
missing data -- `na_action = "mask"` treats a missing point as an
independent observation of variance `mask_variance` (`1` by default),
preserving positive-definiteness:

```{r}
k <- se_kernel(length_scale = 1)
evaluate(k, c(0, NA, 2), na_action = "mask")
```

Compare with the default:

```{r}
evaluate(k, c(0, NA, 2)) # na_action = "propagate" (the default)
```

`mask_variance` is adjustable, e.g. to match the scale of the data's own
noise:

```{r}
evaluate(k, c(0, NA, 2), na_action = "mask", mask_variance = 0.01)
```

`na_action`/`mask_variance` are forwarded automatically through any
composition of kernels:

```{r}
k_composite <- se_kernel(length_scale = 1) + white_noise_kernel(noise = 0.1)
evaluate(k_composite, c(0, NA, 2), na_action = "mask")
```

## Writing your own kernel

Every base kernel constructor follows the same recipe. Here is a minimal
custom stationary kernel (a slightly unusual "cosine" kernel), to show the
pattern end to end:

```{r}
cosine_kernel <- function(length_scale,
                           length_scale_parametrisation = log_exp_parametrisation(),
                           distance_function = euclidean_distance,
                           engine = dense_engine()) {
  kernel <- structure(
    list(
      length_scale = wrapped_param(length_scale, length_scale_parametrisation),
      distance_function = resolve_distance_function(distance_function),
      engine = engine
    ),
    class = c("cosine_kernel", "stationary_kernel", "kernel")
  )
  validate_kernel(kernel)
  kernel
}

# pairwise() defines the formula at the level of a single pair of points --
# the always-correct reference, exercised directly in tests.
pairwise.cosine_kernel <- function(kernel, x1, x2) {
  d <- kernel$distance_function(x1, x2)
  cos(d / unwrap_param(kernel$length_scale))
}

k_cosine <- cosine_kernel(length_scale = 1)
evaluate(k_cosine, c(0, 1, 2))
```

This already works end to end: `evaluate()`, `get_params()`, `kupdate()`,
`freeze()`, printing, and composition via `+`/`*` all work immediately,
because they only depend on the generic `$length_scale`/`$distance_function`
structure, not on anything specific to `cosine_kernel`:

```{r}
k_cosine + white_noise_kernel(noise = 0.1)
kupdate(k_cosine, length_scale = 3)
get_params(freeze(k_cosine, "length_scale"))
```

`vignette("c-expert")` shows how to additionally give a custom stationary
kernel a fast, vectorised `evaluate()` (rather than the always-correct but
slower default), by providing a `shape()` method.

## Where to go next

`vignette("c-expert")`: computation engines (`dense_engine()`,
`diagonal_engine()`), the internal vectorisation machinery, `batch_kernel()`,
`input_specific_param_kernel()`, `block_kernel()`, and `block_diag_kernel()`.
