---
title: "3. Engines, Batching, and Block Structure (Expert)"
subtitle: "Computation engines, vectorisation internals, batching, heteroscedastic noise, and block covariance matrices"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{3. Engines, Batching, and Block Structure (Expert)}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(keRnel)
```

This vignette assumes you have read `vignette("a-beginner")` and
`vignette("b-intermediate")`. It covers the lowest level of the package:
how `evaluate()` actually turns a kernel formula into a matrix, how to make
that fast, and the wrappers that need more than a pointwise transform of
their inner kernel's result (batching, per-input parameters, and block
structure).

## The 4-layer architecture, briefly

Every base kernel's `evaluate()` goes through 4 layers:

1. **`pairwise(kernel, x1, x2)`**: the formula for a *single* pair of
   points -- always correct, the reference definition, exercised directly
   in tests.
2. **A mapping primitive** (`map_pairs()`, `map_diag()`): turns a
   pairwise function into a full result, agnostic of the kernel and the
   engine.
3. **`engine_eval(engine, kernel, X1, X2)`**: picks *which* mapping
   primitive to use -- this is the only layer that knows about the engine.
4. **`evaluate()`**: the user-facing entry point. Coerces inputs, handles
   missing values, applies the memory-size warning, and dispatches.

Composite/wrapper kernels (`sum_kernel()`, `batch_kernel()`,
`block_kernel()`...) skip layers 2-3 entirely: they override `evaluate()`
directly and recombine the `evaluate()` results of their sub-kernels.

## Computation engines

An engine controls *how* a covariance result is computed from a kernel's
`pairwise()` formula -- the policy of *which* mapping primitive to use.
Every kernel has an `engine` field, defaulting to `dense_engine()`.

### `dense_engine()`: the default

Computes the full `N1 x N2` matrix:

```{r}
k <- se_kernel(length_scale = 1, engine = dense_engine())
evaluate(k, c(0, 1, 2))
```

### `diagonal_engine()`: only the diagonal, in `O(n)` instead of `O(n^2)`

When you only need `k(x1[i], x2[i])` for aligned pairs -- e.g. the
predictive variance of a GP -- computing the full dense matrix and then
taking its diagonal wastes `O(n^2)` work. `diagonal_engine()` computes only
what is needed:

```{r}
k_diag <- se_kernel(length_scale = 1, engine = diagonal_engine())
evaluate(k_diag, c(0, 1, 2), c(0, 1, 2))
```

This matches the diagonal of the dense result:

```{r}
identical(
  evaluate(k_diag, c(0, 1, 2), c(0, 1, 2)),
  diag(evaluate(k, c(0, 1, 2), c(0, 1, 2)))
)
```

The performance difference is real and measurable:

```{r}
set.seed(1)
x <- rnorm(2000)

t_dense <- system.time(evaluate(se_kernel(length_scale = 1), x))
t_diag <- system.time(evaluate(se_kernel(length_scale = 1, engine = diagonal_engine()), x, x))

t_dense[["elapsed"]]
t_diag[["elapsed"]]
```

::: {.alert .alert-info}
`diagonal_engine()` is a deliberate design choice: when only the diagonal
of the covariance matrix is needed, computing the full `n1 x n2` matrix and
discarding everything off-diagonal wastes `O(n^2)` work. The optimisation
is purely algebraic, so it pays off regardless of how the underlying
kernel computation is implemented.
:::

`diagonal_engine()` requires `x1` and `x2` to have the same number of
points:

```{r, error = TRUE}
evaluate(k_diag, c(0, 1, 2), c(0, 1))
```

## Vectorisation internals: writing a fast stationary kernel

`vignette("b-intermediate")` showed how to write a minimal custom
stationary kernel with just `pairwise()`. That kernel works correctly, but
`evaluate()` falls back to `map_pairs()` -- an explicit double loop in R,
calling `pairwise()` once per pair of points.

For a stationary kernel, `pairwise(kernel, x1, x2)` is always
`shape(kernel, distance_function(x1, x2))`: a scalar transform of a
distance. Because R vectorises elementwise arithmetic (`exp()`, `^`, ...)
over an entire matrix for free, providing a `shape()` method lets
`evaluate()` apply the kernel's formula to a *whole distance matrix* at
once, skipping the explicit point-by-point loop entirely -- as long as the
distance itself also has a known vectorised form (`distance_matrix()`
already provides one for squared Euclidean, Euclidean, Manhattan, dot
product, and equality; anything else falls back to the same loop, still
correct, just slower):

```{r}
cosine_kernel <- function(length_scale,
                           length_scale_parametrisation = log_exp_parametrisation(),
                           distance_function = euclidean_distance,
                           engine = dense_engine()) {
  kernel <- structure(
    list(
      length_scale = wrapped_param(length_scale, length_scale_parametrisation),
      distance_function = resolve_distance_function(distance_function),
      engine = engine
    ),
    class = c("cosine_kernel", "stationary_kernel", "kernel")
  )
  validate_kernel(kernel)
  kernel
}

pairwise.cosine_kernel <- function(kernel, x1, x2) {
  shape(kernel, kernel$distance_function(x1, x2))
}

# The one extra method that unlocks the fast, vectorised evaluate():
shape.cosine_kernel <- function(kernel, d) {
  cos(d / unwrap_param(kernel$length_scale))
}

k_cosine <- cosine_kernel(length_scale = 1)
evaluate(k_cosine, c(0, 1, 2))
```

Writing `pairwise()` as `shape(kernel, distance)` keeps a single source of
truth for the formula: `pairwise()` remains the always-correct reference
(used directly in tests), while `evaluate()` transparently benefits from
the fast path through `pairwise_matrix.stationary_kernel()`. You can verify
both paths agree, computing the reference the slow way (one pair at a
time, via `outer()`) and comparing it to `evaluate()`'s vectorised result:

```{r}
x <- c(-2, -1, 0, 1, 2)
reference <- outer(x, x, Vectorize(function(a, b) pairwise(k_cosine, a, b)))
identical(evaluate(k_cosine, x), reference)
```

The same pattern (a `shape()` method) applies to a custom
`dot_product_kernel`, with `pairwise_matrix.dot_product_kernel()` reusing
the vectorised dot-product fast path (`X1 %*% t(X2)`) the same way.

## `batch_kernel()`: many covariance matrices at once

Wraps a kernel so `evaluate()` produces several matrices at once, one per
"batch element" -- typically because each batch element has its own
hyperparameter values. Which hyperparameters vary per batch element is an
explicit contract: by default it is auto-detected from the **length** of
the value you give it (length 1 means shared, length `batch_size` means
one value per element -- any other length is an error at construction, not
a silent recycling), and it can always be declared by name via the
`batch_axes` argument when auto-detection would be ambiguous (see
`vignette("d-batching")` for the full story):

```{r}
k_batch <- batch_kernel(se_kernel(length_scale = c(1, 2, 3)), batch_size = 3)
result <- evaluate(k_batch, c(0, 1, 2))
dim(result) # (n1, n2, batch_size) -- batch dimension last, the natural
            # shape produced by simplify2array()
result[, , 2] # exactly se_kernel(length_scale = 2)'s matrix
```

A hyperparameter left as a single scalar is shared across every batch
element:

```{r}
k_batch_shared <- batch_kernel(se_kernel(length_scale = 2), batch_size = 4)
result_shared <- evaluate(k_batch_shared, c(0, 1))
identical(result_shared[, , 1], result_shared[, , 4])
```

`batch_over_inputs = TRUE` additionally lets every batch element see a
*different* set of input points, passed as a 3-dimensional array of shape
`(batch_size, n_points, n_dims)`:

```{r}
x1 <- array(0, dim = c(2, 3, 1))
x1[1, , ] <- c(0, 1, 2)
x1[2, , ] <- c(10, 11, 12)

k_batch_inputs <- batch_kernel(se_kernel(length_scale = 1), batch_size = 2, batch_over_inputs = TRUE)
result_inputs <- evaluate(k_batch_inputs, x1)
result_inputs[, , 1] # se_kernel evaluated on c(0, 1, 2)
result_inputs[, , 2] # se_kernel evaluated on c(10, 11, 12)
```

### Nested batching

A `batch_kernel()` can itself be wrapped inside another `batch_kernel()`,
to handle more than one batch dimension at once. Declare each level's
`batch_axes` explicitly so that every level owns its axis (each level then
slices only its own hyperparameters, and levels may have different sizes
-- see `vignette("d-batching")` for a worked two-hyperparameter grid):

```{r}
k_nested <- batch_kernel(
  batch_kernel(se_kernel(length_scale = c(1, 2)) + white_noise_kernel(noise = c(.1, .2, .3)),
               batch_size = 2, batch_axes = "length_scale"),
  batch_size = 3, batch_axes = "noise"
)
get_params(k_nested) # all values, regardless of nesting depth
dim(evaluate(k_nested, c(0, 1))) # (n, n, inner batch, outer batch)
```

`get_params()`/`kupdate()`/`freeze()` work through any nesting depth
without modification, because they recurse through `$subkernels` uniformly.

::: {.alert .alert-info}
This is more than a cosmetic convenience. `frozen` lives directly on each
`wrapped_param`, with no parallel mask structure that needs to mirror the
kernel's shape. That design avoids an entire class of bugs that a
separate-mask-tree approach would be exposed to once batching nests:
because freezing is a property of the parameter itself rather than of a
side structure that must stay in sync, nested `batch_kernel()`s freeze and
unfreeze correctly with no special-casing.
:::

## `input_specific_param_kernel()`: a parameter per input point

Similar to `batch_kernel()`, but varies a hyperparameter per **point of
`x1`** rather than per batch element -- the canonical use case is
heteroscedastic noise (a different noise level for each observation):

```{r}
k_het_noise <- input_specific_param_kernel(
  white_noise_kernel(noise = c(0.1, 0.5, 0.2)),
  input_size = 3
)
evaluate(k_het_noise, c(0, 1, 2))
```

Combine with `+` for a GP with heteroscedastic observation noise on top of
a smooth signal kernel:

```{r}
k_model <- se_kernel(length_scale = 1) + k_het_noise
diag(evaluate(k_model, c(0, 1, 2)))
```

## Block covariance matrices

### `block_diag_kernel()`: independent blocks

The simple case: builds a block-diagonal matrix of `nb_blocks` *independent*
blocks (off-diagonal blocks are exactly `0`). Internally, this is just
`batch_kernel()` plus reassembling the result:

```{r}
k_block_diag <- block_diag_kernel(se_kernel(length_scale = c(1, 2)), nb_blocks = 2)
evaluate(k_block_diag, c(0, 1))
```

### `block_kernel()`: blocks with cross-terms

The harder case: also computes the *cross*-block terms (block `(i, j)` for
`i != j`), rather than leaving them at `0`. This only makes mathematical
sense if the inner kernel can read a **pair** of hyperparameter values at
once (one per "side" of the block) -- most kernels cannot.
`feature_kernel()` is designed for exactly this:

```{r}
inner <- feature_kernel(length_scale = c(1, 3), length_scale_u = 1, variance = c(2, 1))
k_block <- block_kernel(inner, nb_blocks = 2)
evaluate(k_block, c(0, 1))
```

A kernel that does *not* know how to interpret a paired hyperparameter
value raises an explicit error rather than silently producing a
mathematically meaningless result:

```{r, error = TRUE}
block_kernel(se_kernel(length_scale = c(1, 2)), nb_blocks = 2)
```

...unless every hyperparameter is shared (length 1), in which case no
pairing is ever attempted and every block is simply identical:

```{r}
k_block_shared <- block_kernel(se_kernel(length_scale = 1), nb_blocks = 2)
result <- evaluate(k_block_shared, c(0, 1))
identical(result[1:2, 1:2], result[3:4, 3:4])
```

Internally, `block_kernel()` only computes the upper triangle of blocks and
mirrors the rest by transposition, exploiting the fact that a
`block_compatible` kernel's formula is symmetric under swapping which side
is "left"/"right" together with which input is `x1`/`x2`:

```{r}
identical(evaluate(k_block, c(0, 1))[1:2, 3:4], t(evaluate(k_block, c(0, 1))[3:4, 1:2]))
```

A kernel must be explicitly marked block-compatible (an attribute,
`attr(kernel, "block_compatible")`) to be used this way -- `feature_kernel()`
already carries this marker:

```{r}
isTRUE(attr(feature_kernel(length_scale = 1, length_scale_u = 1, variance = 1), "block_compatible"))
```

## The memory-size warning guard

Building a dense `n1 x n2` matrix gets expensive quickly, and R has neither
JIT compilation nor GPU offload to absorb the cost. `evaluate()` warns (not
errors) above a configurable threshold:

```{r}
withCallingHandlers(
  evaluate(se_kernel(length_scale = 1), seq_len(10), memory_warning_threshold = 5),
  warning = function(w) {
    message("Caught: ", conditionMessage(w))
    invokeRestart("muffleWarning")
  }
)
```

Raise the threshold (or set it to `Inf`) to silence it once you know what
you are doing:

```{r}
result <- evaluate(se_kernel(length_scale = 1), seq_len(10), memory_warning_threshold = Inf)
dim(result)
```

## Summary

This vignette covered the more advanced machinery `keRnel` provides for
building and composing kernels efficiently:

- `dense_engine()` and `diagonal_engine()` let you choose how a covariance
  matrix is computed, picking the cheaper diagonal-only path when that's
  all you need.
- `batch_kernel()` produces several covariance matrices at once, with
  per-batch-element hyperparameters inferred automatically from value
  length, and nests cleanly without special-casing.
- `input_specific_param_kernel()` varies a hyperparameter per input point,
  the natural fit for heteroscedastic noise.
- `block_kernel()`/`block_diag_kernel()` build block covariance matrices
  for grouped data, relying on the `block_compatible` marker.
- The memory-size warning guard in `evaluate()` flags expensive dense
  matrix builds before they happen, without forcing you to abandon them.
- `kernel_interactions()` is a standalone ergonomic feature for inspecting
  how composed kernels relate to their subkernels.
