---
title: "4. Batching: one kernel, many configurations"
subtitle: "Batched hyperparameters, batched inputs, nested batches, per-point noise, and what they mean for a Gaussian Process"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{4. Batching: one kernel, many configurations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 4.5
)
set.seed(42)
```

```{r setup}
library(keRnel)
```

This vignette assumes you are comfortable with the material of the beginner
and intermediate vignettes (`vignette("a-beginner")`,
`vignette("b-intermediate")`): building kernels, `evaluate()`,
`get_params()`/`kupdate()`/`freeze()`.

## Why batching? One prior is rarely enough

A Gaussian Process prior is entirely determined by its kernel and the
kernel's hyperparameters:

$$f \sim \mathcal{GP}\big(0,\; k_\theta(x, x')\big)$$

In practice you almost never work with a *single* $\theta$. You compare a
grid of candidate length-scales, you restart an optimiser from several
initial values, you fit the same model to several subjects, you give every
observation its own noise level. All of these are the same computational
pattern: **evaluate one kernel under a family of configurations**
$\{\theta_1, \ldots, \theta_B\}$, against shared or per-configuration
inputs.

You could write that as a loop. `batch_kernel()` instead makes the family
itself a kernel object: build the inner kernel with a *vector* of
hyperparameter values (one per batch element) and `evaluate()` returns all
$B$ covariance matrices at once, stacked along a trailing dimension.

## Batching over hyperparameters: what does a length-scale do?

The squared-exponential kernel is

$$k_\ell(x, x') = \exp\!\left(-\frac{(x - x')^2}{2\ell^2}\right),$$

where the length-scale $\ell$ controls how far apart two points can be
while remaining correlated. To *see* what $\ell$ does, evaluate the same
kernel under three values of it at once:

```{r}
x <- seq(0, 10, length.out = 200)

k <- batch_kernel(
  se_kernel(length_scale = c(0.3, 1, 3)),
  batch_size = 3
)

K <- evaluate(k, x, memory_warning_threshold = Inf)
dim(K)
```

The result is an array of shape `(n, n, batch_size)`: `K[, , b]` is the
covariance matrix of the GP prior with the `b`-th length-scale. A draw from
that prior is $f = L z$ with $L L^\top = K$ and $z \sim \mathcal{N}(0, I)$:

```{r}
sample_gp_prior <- function(K, n_draws = 2) {
  L <- t(chol(K + diag(1e-9, nrow(K)))) # jitter for numerical stability
  L %*% matrix(rnorm(nrow(K) * n_draws), nrow(K), n_draws)
}

ls_values <- c(0.3, 1, 3)
draws <- lapply(1:3, function(b) sample_gp_prior(K[, , b]))

matplot(x, do.call(cbind, draws),
        type = "l", lty = rep(1:2, times = 3), lwd = 1.5,
        col = rep(c("#D55E00", "#0072B2", "#009E73"), each = 2),
        xlab = "x", ylab = "f(x)",
        main = "GP prior draws under three length-scales, one batched kernel")
legend("topright", bty = "n", lwd = 1.5,
       col = c("#D55E00", "#0072B2", "#009E73"),
       legend = paste("length_scale =", ls_values))
```

Small $\ell$ wiggles fast, large $\ell$ produces slowly-varying, almost
linear functions -- and all three covariance matrices came from a *single*
`evaluate()` call on a single kernel object.

A hyperparameter left as a single scalar is shared across every batch
element. This is how you batch only *part* of a kernel:

```{r}
# Three noise levels, one shared length-scale:
k_mixed <- batch_kernel(
  se_kernel(length_scale = 1) + white_noise_kernel(noise = c(0.01, 0.1, 1)),
  batch_size = 3
)
K_mixed <- evaluate(k_mixed, c(0, 1, 2))
# The diagonal (signal variance 1 + noise) changes; off-diagonals don't:
sapply(1:3, function(b) K_mixed[1, 1, b])
sapply(1:3, function(b) K_mixed[1, 2, b])
```

In GP terms this is the *signal-plus-noise* decomposition
$k(x, x') = k_{\mathrm{SE}}(x, x') + \sigma_b^2 \, \delta_{x = x'}$ with a
per-configuration noise variance $\sigma_b^2$ -- the exact structure you
sweep when profiling a noise level.

## Batching over inputs: same prior, several datasets

The dual situation: one set of hyperparameters, but $B$ different input
sets -- say, $B$ subjects observed at different times, all sharing the same
GP prior. Pass `batch_over_inputs = TRUE` and give `evaluate()` a 3-D array
of shape `(batch_size, n_points, n_dims)`:

```{r}
k_shared <- batch_kernel(se_kernel(length_scale = 1),
                         batch_size = 2, batch_over_inputs = TRUE)

x_subjects <- array(0, dim = c(2, 4, 1))
x_subjects[1, , 1] <- c(0.0, 0.5, 1.0, 1.5)  # subject 1's design
x_subjects[2, , 1] <- c(0.0, 2.0, 4.0, 6.0)  # subject 2's design

K_subjects <- evaluate(k_shared, x_subjects)
dim(K_subjects)
# Subject 2's points are further apart => weaker correlations:
K_subjects[1, 2, 1] # subjects 1: k(0, 0.5)
K_subjects[1, 2, 2] # subjects 2: k(0, 2)
```

Both flavours combine freely: batched inputs *and* per-batch
hyperparameters give each subject its own design and its own
hyperparameter values -- a fully independent-subjects model evaluated in
one call.

## The axis contract: declaring what varies

So far, `batch_kernel()` *auto-detected* which hyperparameter was batched:
any multi-valued hyperparameter is assumed to vary along the batch axis,
and must then have exactly `batch_size` values. The validation is strict on
purpose -- a mismatched length is an error at construction, never a silent
recycling:

```{r, error = TRUE}
batch_kernel(se_kernel(length_scale = c(1, 2, 3)), batch_size = 5)
```

But "this vector has several values" does not always mean "one value per
batch element". The clearest counter-example is [ard_kernel()], whose
`length_scales` has one value per *input dimension*:

$$k_{\mathrm{ARD}}(x, x') = k\!\left(\frac{x_1}{\ell_1}, \ldots, \frac{x_D}{\ell_D};\;
\frac{x'_1}{\ell_1}, \ldots, \frac{x'_D}{\ell_D}\right)$$

Those $D$ values index dimensions, never batch elements -- even when $D$
happens to equal `batch_size`. `keRnel` encodes that distinction on the
parameter itself (`wrapped_param(..., sliceable = FALSE)`), so an ARD
kernel inside a batch is never mis-sliced, coincidence or not:

```{r}
ard <- ard_kernel(se_kernel, length_scales = c(1, 2, 3)) # 3 input dimensions
bk <- batch_kernel(ard, batch_size = 3)                  # batch of 3: no clash

X <- matrix(rnorm(12), ncol = 3)
K_ard <- evaluate(bk, X)
# Nothing was batched, so all three matrices are the full ARD kernel:
all.equal(K_ard[, , 1], evaluate(ard, X))
```

When auto-detection *is* ambiguous, name the batched hyperparameters
explicitly with `batch_axes`. A multi-valued hyperparameter **not** named
there is left alone -- it belongs to some other wrapper's axis. That rule
is what makes nested batching work.

## Nested batching: a grid over two hyperparameters

Suppose you want to explore a full grid: 2 candidate length-scales
$\times$ 4 candidate noise levels, i.e. the 8 kernels

$$k_{i,o}(x, x') = \exp\!\left(-\frac{(x-x')^2}{2\ell_i^2}\right)
                  + \sigma_o^2\, \delta_{x = x'},
  \qquad i \in \{1, 2\},\; o \in \{1, \ldots, 4\}.$$

Wrap a `batch_kernel()` inside another one, and give each level its own
axis:

```{r}
k <- se_kernel(length_scale = c(0.5, 1.5)) +
     white_noise_kernel(noise = c(0.5, 1, 2, 3))

inner <- batch_kernel(k, batch_size = 2, batch_axes = "length_scale")
grid  <- batch_kernel(inner, batch_size = 4, batch_axes = "noise")

K_grid <- evaluate(grid, c(1, 2, 3))
dim(K_grid) # (n, n, inner batch, outer batch)
```

`K_grid[, , i, o]` is exactly $k_{i,o}$'s covariance matrix. Two structural
checks make the "each level owns its axis" guarantee visible -- the
diagonal only responds to the *outer* (noise) axis, the off-diagonal only
to the *inner* (length-scale) one:

```{r}
# Diagonal = 1 + noise, whatever the length-scale:
sapply(1:4, function(o) K_grid[1, 1, 1, o])
sapply(1:4, function(o) K_grid[1, 1, 2, o])

# Off-diagonal = exp(-d^2 / (2 l^2)), whatever the noise:
K_grid[1, 2, 1, 1] - exp(-2)     # l = 0.5, d = 1
K_grid[1, 2, 2, 1] - exp(-2 / 9) # l = 1.5, d = 1
```

(Readers coming from the Python sibling library `kernax` will recognise
this example: it is the same two-level batch as its
`test_double_batch_with_masks`, with the same reference values -- the two
libraries produce the same grid, modulo R's convention of putting batch
dimensions last.)

If you forget to declare the axes, nothing silently breaks -- construction
fails loudly, telling you which lengths clashed and to use `batch_axes`:

```{r, error = TRUE}
batch_kernel(k, batch_size = 2)
```

## Per-point hyperparameters: heteroscedastic noise

Batching varies a hyperparameter *across covariance matrices*.
[input_specific_param_kernel()] varies one *within* a single matrix, along
the points of `x1`. Its canonical use is heteroscedastic regression, where
observation $i$ has its own noise variance $\sigma_i^2$:

$$y_i = f(x_i) + \varepsilon_i, \qquad
  \varepsilon_i \sim \mathcal{N}(0, \sigma_i^2)
  \quad\Longrightarrow\quad
  \mathbf{K}_y = \mathbf{K}_f + \mathrm{diag}(\sigma_1^2, \ldots, \sigma_n^2).$$

```{r}
x_obs <- c(1, 3, 5, 7, 9)
sigma2 <- c(0.01, 0.01, 1.5, 0.01, 0.01) # the third point is much noisier

k_y <- se_kernel(length_scale = 1.2) +
  input_specific_param_kernel(white_noise_kernel(noise = sigma2),
                              input_size = 5)

K_y <- evaluate(k_y, x_obs)
diag(K_y) # 1 + sigma_i^2, point by point
```

The GP posterior mean at new points $x_*$ is
$\mu_* = \mathbf{K}_*^\top \mathbf{K}_y^{-1} y$, and the noisy point simply
counts for less:

```{r}
y_obs <- sin(x_obs)
x_new <- seq(0, 10, length.out = 100)

k_f <- se_kernel(length_scale = 1.2)
K_star <- evaluate(k_f, x_obs, x_new)
K_new_diag <- rep(1, length(x_new)) # k(x, x) = 1 for the SE kernel

alpha <- solve(K_y, y_obs)
post_mean <- drop(crossprod(K_star, alpha))
post_var <- pmax(K_new_diag - colSums(K_star * solve(K_y, K_star)), 0)

plot(x_new, post_mean, type = "l", lwd = 2, col = "#0072B2",
     ylim = range(post_mean + 2 * sqrt(post_var),
                  post_mean - 2 * sqrt(post_var), y_obs),
     xlab = "x", ylab = "f(x)",
     main = "Heteroscedastic GP: the noisy point (x = 5) is trusted less")
polygon(c(x_new, rev(x_new)),
        c(post_mean + 2 * sqrt(post_var), rev(post_mean - 2 * sqrt(post_var))),
        col = adjustcolor("#0072B2", alpha.f = 0.15), border = NA)
points(x_obs, y_obs, pch = 19,
       col = ifelse(sigma2 > 1, "#D55E00", "black"), cex = 1.2)
```

Note the posterior mean does not bend to pass through the orange (noisy)
point, and the uncertainty band barely narrows there.

**A caveat that matters mathematically**: entry $(i, j)$ of this wrapper's
auto-covariance uses point $i$'s value and entry $(j, i)$ uses point
$j$'s. The result is only a symmetric, valid covariance matrix when the
varied hyperparameter acts on the *diagonal only* -- which is exactly why
`white_noise_kernel()`'s `noise` is its canonical use. Varying, say, a
`length_scale` per point would produce an asymmetric matrix: no longer a
GP covariance.

## Missing data across structured kernels

The intermediate vignette introduced `na_action = "mask"`: a point with a
missing value is treated as an independent pseudo-observation of variance
`mask_variance`, keeping $\mathbf{K}$ invertible. That promise requires
knowing *which entries are the self-variances of the masked point* -- a
question that only makes sense for the matrix the **user** asked for, not
for the internal slices a structured kernel evaluates along the way (a
single row of `x1`, one block of a block matrix...).

`keRnel` therefore decides "is this a self-covariance?" once, at the level
of your call, and forwards the answer through every internal
re-evaluation. The result: masking behaves identically however deeply the
kernel is structured.

```{r}
x_na <- c(0, NA, 2)

# Through a per-point-noise wrapper: the masked point keeps its unit
# self-variance, the matrix stays invertible...
k_het <- input_specific_param_kernel(white_noise_kernel(noise = c(.1, .5, .2)),
                                     input_size = 3)
K_masked <- evaluate(k_het, x_na, na_action = "mask")
K_masked
chol(K_masked)[1, 1] # factorisation succeeds: the matrix is usable
```

```{r}
# ...and through a batch, every batch element gets the same treatment:
k_b <- batch_kernel(se_kernel(length_scale = c(1, 2)), batch_size = 2)
K_b <- evaluate(k_b, x_na, na_action = "mask")
K_b[, , 1]
```

For a *block* kernel (below), there is one more subtlety, handled for you:
the diagonal blocks of a self-covariance are self-covariances, but a
*cross* block covaries two different processes -- there, a masked point
contributes `0`, never a self-variance.

## Multi-output GPs: block kernels

A two-output GP stacks outputs into one covariance matrix of blocks:

$$\mathbf{K} = \begin{pmatrix}
  \mathbf{K}_{11} & \mathbf{K}_{12} \\
  \mathbf{K}_{12}^\top & \mathbf{K}_{22}
\end{pmatrix},$$

where $\mathbf{K}_{bb'}$ covaries output $b$ with output $b'$. Computing a
cross block needs *both* blocks' hyperparameter values at once, so the
inner kernel must be designed for pairs -- [feature_kernel()] is the
built-in prototype (kernels not marked block-compatible are rejected
rather than silently misused):

```{r}
inner <- feature_kernel(length_scale = c(1, 3), length_scale_u = 1,
                        variance = c(2, 1))
k_mo <- block_kernel(inner, nb_blocks = 2)
K_mo <- evaluate(k_mo, c(0, 1, 2))
round(K_mo, 3)
```

The off-diagonal $3 \times 3$ blocks are the cross-covariances between the
two outputs: non-zero (the outputs are correlated), and the transpose of
each other (the matrix stays symmetric). With `na_action = "mask"`, a
missing point receives `mask_variance` in each *diagonal* block only:

```{r}
K_mo_na <- evaluate(k_mo, c(0, NA, 2), na_action = "mask")
K_mo_na[2, c(2, 5)] # own-block diagonal: masked; cross block: 0
```

## Batching and training configurations

Batching composes with the training toolkit from the intermediate
vignette, because slicing only ever touches values -- `frozen` flags and
parametrisations travel with each slice untouched:

```{r}
k_train <- batch_kernel(
  se_kernel(length_scale = c(0.5, 1, 2)) + white_noise_kernel(noise = 0.1),
  batch_size = 3
)
k_train <- freeze(k_train, "noise")

get_trainable_params(k_train) # the 3 batched length-scales; noise excluded
```

A typical multi-start workflow draws candidate starting values with
`sample_hps()`, packs them into one batched kernel, and evaluates every
candidate's covariance in a single call -- the batched analogue of
restarting an optimiser $B$ times.

Finally, the axis contract is re-checked at every `evaluate()`, so a
later `kupdate()` cannot silently desynchronise a batch:

```{r, error = TRUE}
k_drift <- kupdate(k_train, length_scale = c(1, 2)) # 2 values, batch of 3
evaluate(k_drift, c(0, 1))
```

## Where to go next

- `?batch_kernel`, `?input_specific_param_kernel`, `?block_kernel` for the
  full contracts (including `batch_axes`/`param_axes`/`block_axes`).
- `vignette("c-expert")` for writing your own kernels -- any custom kernel
  built from `wrapped_param()`s gets batching support for free.
