---
title: "Advanced usage"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Advanced usage}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```





## Table of Contents
- [1. Introduction](#introduction)
- [2. Commonly Used Kernels](#commonly-used-kernels)
  - [2.1 Squared Exponential (SE)](#squared-exponential-se)
  - [2.2 Rational Quadratic (RQ)](#rational-quadratic-rq)
  - [2.3 Periodic](#periodic)
  - [2.4 Matern](#matern)
  - [2.5 Linear](#linear)
  - [2.6 Constant](#constant)
  - [2.7 Noise](#noise)
- [3. Combining Kernels](#combining-kernels)
- [4. Custom Kernels](#custom-kernels)

## 1. Introduction {#introduction}


In this vignette, we explore the diversity of *kernels* available in the **keRnel** package. We begin by examining commonly used kernels, then demonstrate how to combine them to address specific problems. Finally, we show how to extend the `AbstractKernel` class to create custom kernels.


 

```{r setup}
library(keRnel)
```

## 2. Commonly used kernels already implemented {#commonly-used-kernels}

This section provides an overview of the kernels available in the **keRnel** package. For detailed documentation, refer to:

```{r eval=FALSE}
?kernel
```





| Kernel Type         | Class Name           | Key Parameters                          |
|---------------------|----------------------|------------------------------------------|
| Squared Exponential | SEKernel             | variance_se, length_scale_se             |
| Rational Quadratic  | RationalQuadraticKernel | variance_rq, length_scale_rq, alpha_rq |
| Periodic            | PeriodicKernel       | variance_per, length_scale_per, period  |
| Matern 1/2          | MaternKernel12       | length_scale_mat                        |
| Matern 3/2          | MaternKernel32       | length_scale_mat                        |
| Matern 5/2          | MaternKernel52       | length_scale_mat                        |
| Linear              | LinearKernel         | sigma2_b, sigma2_v, c                   |
| Constant            | ConstantKernel       | value_c                                  |
| Noise               | NoiseKernel          | value_c                                  |

```{r}
#Input for visualization 
x = matrix(rnorm(100), ncol = 2)

```


### 2.1 Squared Exponential {#squared-exponential-se}

The **Squared Exponential (SE) kernel** is ideal for modeling **smooth functions** with infinite differentiability.

#### Definition
Let \( \sigma \in \mathbb{R}_{+} \) and \( \ell \in \mathbb{R}_{+}^{*} \). Consider the squared exponential kernel defined by:

\[
K_{\text{SE}}(x, x') = \sigma^2 \exp\left(-\frac{\|x - x'\|^2}{2\ell^2}\right)
\]

where $\sigma$ is the signal standard deviation and $\ell$ is the length scale parameter.

```{r}
# Create an object of the SEKernel class.
se_kernel = new("SEKernel")
pretty_print(se_kernel)
```

It is possible to access both hyperparameters (${\sigma}$ and${\ell}$) and modify them.

```{r}
get_hyperparameter_values(se_kernel)

hp = c(12,1)
se_kernel = set_hyperparameters(se_kernel, hp)
pretty_print(se_kernel)
```
#### Visualization 

One can observe the heatmap of the covariance matrix obtained using this kernel.
```{r warning=FALSE}
visualize_kernel(se_kernel, x)
```

We can observe the relationship between the distance between the inputs and the kernel.
```{r}
plot_kernel_vs_distance(se_kernel, x)
```



### 2.2. Rational Quadratic {#rational-quadratic-rq}
**Rational Quadratic kernel** is suitable for functions with multiple length scales.
#### Definition

Let \( \sigma^2 \in \mathbb{R}_{+} \), \( \alpha \in \mathbb{R}_{+}^{*} \), and \( \ell \in \mathbb{R}_{+}^{*} \). Consider the RQ kernel defined by:

\[
K_{\text{RQ}}(x, x')  = \sigma^2 \left(1 + \frac{\|x - x'\|^2}{2 \alpha \ell^2}\right)^{-\alpha}
\]


```{r}
# Create an object of the SEKernel class.
RQ_kernel = new("RationalQuadraticKernel")
pretty_print(RQ_kernel)
```

It is possible to access both hyperparameters (${\sigma}$ and${\ell}$) and modify them.

```{r}
get_hyperparameter_values(RQ_kernel)

hp = c(12,1,5)
RQ_kernel = set_hyperparameters(RQ_kernel, hp)
pretty_print(RQ_kernel)
```
#### Visualization 

One can observe the heatmap of the covariance matrix obtained using this kernel.
```{r warning=FALSE}
visualize_kernel(RQ_kernel, x)
```

We can observe the relationship between the distance between the inputs and the kernel.
```{r}
plot_kernel_vs_distance(RQ_kernel, x)
```






### 2.3. Periodic Kernel {#periodic}

**Periodic Kernel** captures recurrent patterns (e.g., seasonal sales, circadian rhythms).
#### Definition

Let \( \sigma^2 \in \mathbb{R}_{+} \), \( \ell \in \mathbb{R} \), and \( p \in \mathbb{R} \). Consider the periodic kernel defined by:

\[
K_{\text{Per}}(x, x') = \sigma^2 \exp \left( -\frac{2}{\ell^2} \sin^2 \left( \pi\frac{ x - x'}{p} \right) \right)
\]


```{r}
# Create an object of the SEKernel class.
per_kernel = new("PeriodicKernel")
pretty_print(per_kernel)
```

It is possible to access both hyperparameters (${\sigma}$ and${\ell}$) and modify them.

```{r}
get_hyperparameter_values(per_kernel)

hp = c(12,1,5)
per_kernel = set_hyperparameters(per_kernel, hp)
pretty_print(per_kernel)
```
#### Visualization 

One can observe the heatmap of the covariance matrix obtained using this kernel.
```{r warning=FALSE}
visualize_kernel(per_kernel, x)
```

We can observe the relationship between the distance between the inputs and the kernel.
```{r}
plot_kernel_vs_distance(per_kernel, x)
```





### 2.4. Matern Kernel {#matern}

Here we implement three common versions of the Matern kernel.

#### Definition with $\nu = 1/2 $

**Matern Kernel with $\nu = 1/2 $** models non-smooth functions (e.g., Wi-Fi signal strength, rough terrain).


Let \( \ell \in \mathbb{R}_{+}^{*}, \sigma^2 \in \mathbb{R}_{+} \). Consider the Matern Kernel with \(\nu\) =1/2 defined by:

\[
K_{\text{Mat}_{\nu = \frac{1}{2}}}(x, x') =  \sigma^2\cdot exp\left(-\frac{\|x - x'\|^2}{\ell}\right)
\]



```{r}
# Create an object of the SEKernel class.
mat_kernel12 = new("MaternKernel12")
pretty_print(mat_kernel12)
```

It is possible to access both hyperparameters (${\sigma}$ and${\ell}$) and modify them.

```{r}
get_hyperparameter_values(mat_kernel12)

hp = c(12)
mat_kernel12 = set_hyperparameters(mat_kernel12, hp)
pretty_print(mat_kernel12)
```
#### Visualization 

One can observe the heatmap of the covariance matrix obtained using this kernel.
```{r warning=FALSE}
visualize_kernel(mat_kernel12, x)
```

We can observe the relationship between the distance between the inputs and the kernel.
```{r}
plot_kernel_vs_distance(mat_kernel12, x)
```


#### Definition with $\nu = 3/2 $

**Matern kernel with  $\nu = 3/2 $** balances smoothness and flexibility.


Let \( \ell \in \mathbb{R}_{+}^{*} \), \(\sigma^2 \in \mathbb{R}_{+}\). Consider the Matern Kernel with \(\nu\) =3/2 defined by:

\[
K_{\text{Mat}_{\nu = \frac{3}{2}}}(x, x') =   \sigma^2\cdot\left( \frac{\sqrt{3}\cdot\|x - x'\|^2}{\ell} +1\right)exp\left(-\frac{\sqrt{3}\cdot\|x - x'\|^2}{\ell}\right)
\]



```{r}
# Create an object of the SEKernel class.
mat_kernel32 = new("MaternKernel32")
pretty_print(mat_kernel32)
```

It is possible to access both hyperparameters (${\sigma}$ and${\ell}$) and modify them.

```{r}
get_hyperparameter_values(mat_kernel32)

hp = c(12)
mat_kernel32 = set_hyperparameters(mat_kernel32, hp)
pretty_print(mat_kernel32)
```
#### Visualization 

One can observe the heatmap of the covariance matrix obtained using this kernel.
```{r warning=FALSE}
visualize_kernel(mat_kernel32, x)
```

We can observe the relationship between the distance between the inputs and the kernel.
```{r}
plot_kernel_vs_distance(mat_kernel32, x)
```


#### Definition avec $\nu = 5/2 $

**Matern kernel with  $\nu = 3/2 $** is almost as smooth as SE but with faster decay.


Let \( \ell \in \mathbb{R}_{+}^{*} \), \( \sigma^2 \in \mathbb{R}_{+} \). Consider the Matern Kernel with \(\nu\) =5/2 defined by:

\[
K_{\text{Mat}_{\nu = \frac{5}{2}}}(x, x') =   \sigma^2\cdot\left( \left( \frac{5(\|x - x'\|^2)^2}{3\ell^2} + \frac{\sqrt{5}\|x - x'\|^2}{\ell} + 1 \right) \exp \left( -\frac{\sqrt{5}\|x - x'\|^2}{\ell} \right) \right)
\]


```{r}
# Create an object of the SEKernel class.
mat_kernel52 = new("MaternKernel52")
pretty_print(mat_kernel52)
```

It is possible to access both hyperparameters (${\sigma}$ and${\ell}$) and modify them.

```{r}
get_hyperparameter_values(mat_kernel52)

hp = c(12)
mat_kernel52 = set_hyperparameters(mat_kernel52, hp)
pretty_print(mat_kernel52)
```
#### Visualization 

One can observe the heatmap of the covariance matrix obtained using this kernel.
```{r warning=FALSE}
visualize_kernel(mat_kernel52, x)
```

We can observe the relationship between the distance between the inputs and the kernel.
```{r}
plot_kernel_vs_distance(mat_kernel52, x)
```



### 2.5. Linear Kernel  {#linear}

**Linear Kernel** is used for linear relationships (e.g., polynomial regression, feature interactions). 

#### Definition

Let \( \sigma^2_b \in \mathbb{R}_{+} \), \( \sigma^2_v \in \mathbb{R}_{+} \), and \( c \in \mathbb{R} \). Consider the linear kernel defined by:

\[
K_{\text{Lin}}(x, x') = \sigma^2_b + \sigma^2_v (x - c)(x' - c)
\]


```{r}
# Create an object of the SEKernel class.
lin_kernel = new("LinearKernel")
pretty_print(lin_kernel)
```

It is possible to access both hyperparameters (${\sigma}$ and${\ell}$) and modify them.

```{r}
get_hyperparameter_values(lin_kernel)

hp = c(12,1,5)
lin_kernel = set_hyperparameters(lin_kernel, hp)
pretty_print(lin_kernel)
```
#### Visualization 

One can observe the heatmap of the covariance matrix obtained using this kernel.
```{r warning=FALSE}
visualize_kernel(lin_kernel, x)
```

We can observe the relationship between the distance between the inputs and the kernel.
```{r}
plot_kernel_vs_distance(lin_kernel, x)
```


### 2.6. Constant Kernel {#constant}

**Constant Kernel**  adds,for instance,  a global bias.

#### Definition

Let \( c \in \mathbb{R}_{+}^{*} \). Consider the constant kernel defined by:

\[
K_{\text{cst}}(x, x') = c \cdot \begin{bmatrix}
1 & 1 & \cdots & 1 \\
1 & 1 & \cdots & 1 \\
\vdots & \vdots & \ddots & \vdots \\
1 & 1 & \cdots & 1
\end{bmatrix}
\]

```{r}
# Create an object of the SEKernel class.
cst_kernel = new("ConstantKernel")
pretty_print(cst_kernel)
```

It is possible to access both hyperparameters (${\sigma}$ and${\ell}$) and modify them.

```{r}
get_hyperparameter_values(cst_kernel)

hp = c(12)
cst_kernel = set_hyperparameters(cst_kernel, hp)
pretty_print(cst_kernel)
```
#### Visualization 

One can observe the heatmap of the covariance matrix obtained using this kernel.
```{r warning=FALSE}
visualize_kernel(cst_kernel, x)
```

We can observe the relationship between the distance between the inputs and the kernel.
```{r}
plot_kernel_vs_distance(cst_kernel, x)
```


### 2.7. Noise Kernel {#noise}
**Noise Kernel** models,for instance, observation noise (e.g., measurement errors in experiments).
#### Definition

Let \( c \in \mathbb{R}_{+}^{*} \). Consider the noise kernel defined by:

\[
K_{\text{noise}}(x, x') = c \cdot I_n
\]

```{r}
# Create an object of the SEKernel class.
noise_kernel = new("NoiseKernel")
pretty_print(noise_kernel)
```

It is possible to access both hyperparameters (${\sigma}$ and${\ell}$) and modify them.

```{r}
get_hyperparameter_values(noise_kernel)

hp = c(12)
noise_kernel = set_hyperparameters(noise_kernel, hp)
pretty_print(noise_kernel)
```
#### Visualization 

One can observe the heatmap of the covariance matrix obtained using this kernel.
```{r warning=FALSE}
visualize_kernel(noise_kernel, x)
```

We can observe the relationship between the distance between the inputs and the kernel.
```{r}
plot_kernel_vs_distance(noise_kernel, x)
```



For detailed documentation, use:

```{r eval=FALSE}
# Access the documentation for a specific kernel
?SEKernel
?RationalQuadraticKernel
# Or browse all available kernels
help(package = "keRnel")
```


## 3. Sum and product of kernels {#combining-kernels}

We can exploit the properties that stipulate that a sum of kernels remains a kernel and that the product of kernels remains a kernel to construct more complex kernels. 

The class system greatly simplifies this part. Objects created by sum or product are elements of the **SumKernel** or **ProductKernel** class, which inherits from **AbstractKernel** like the usual kernels. This means you can access the same title with hyperparameters and use the visualization functions.

### Kernel for periodic data with smooth global variation

```{r}
linear_k = new("LinearKernel",sigma2_b = 1, sigma2_v = 1, c = 0)
noise_k = new("NoiseKernel",value_c = 0.1)

combined_k = linear_k + noise_k

pretty_print(combined_k)
```

```{r warning=FALSE}
visualize_kernel(combined_k, x)
```

```{r}
plot_kernel_vs_distance(combined_k, x)
```



### Kernel spatio-temporal

```{r}
se_space_k = new("SEKernel",variance_se = 1, length_scale_se = 1)  # Space
periodic_time_k = new("PeriodicKernel",variance_per = 1, length_scale_per = 0.5, period = 1)  # Time

combined_k  = se_space_k * periodic_time_k

pretty_print(combined_k)
```


```{r warning=FALSE}
visualize_kernel(combined_k, x)
```

```{r}
plot_kernel_vs_distance(combined_k, x)
```


## 4. Custom Kernels {#custom-kernels}

 
You can extend the AbstractKernel class to implement custom kernels. Here is a minimal example with a kernel based on Minkowski distance :

```{r}
# Define a custom kernel class
setClass(
  "MinkowskiKernel",
  contains = "AbstractKernel",
  slots = c(
    p = "numeric",    # Minkowski distance shape parameter
    sigma = "numeric" # lengthscale
  )
)

# Define the pairwise kernel method
setMethod(
  "pairwise_kernel", "MinkowskiKernel",
  function(obj, x, y) {
    dx = outer(rowSums(x^2), rowSums(y^2), FUN = "+") - 2 * tcrossprod(x, y)
    exp(- dx^obj@p/ obj@sigma)
    
  }
)

# Define the hyperparameters method
setMethod(
  "gt_HPs", "MinkowskiKernel",
  function(obj) {
    list(sigma = obj@sigma, p = obj@p)
  }
)

# Define the pretty print method
setMethod(
  "pretty_print", "MinkowskiKernel",
  function(obj) {
    sprintf("MinkowskiKernel(p=%.2f, sigma=%.2f)", obj@p, obj@sigma)
  }
)

# Créer une instance du noyau
minkowski_kernel = new("MinkowskiKernel", p = 2, sigma = 1.0)
pretty_print(minkowski_kernel)

# Compute the kernel matrix
K_custom = pairwise_kernel(minkowski_kernel, x, x)

print(K_custom[12,3])
```


```{r warning=FALSE}
visualize_kernel(minkowski_kernel, x)
```

```{r}
plot_kernel_vs_distance(minkowski_kernel, x)
```



